Evaluation of Machine Translation Pre-Trained Models for Many-to-English Translation

### 2024-08-20 paper/paper.tex
updated

\title{Evaluation of Pre-Trained Models for Many-to-English Translation}
\author{Stefan Liemawan Adji}
\date{\today}

@@

\begin{abstract}
    Machine translation (MT) has become an essential tool for overcoming language barriers in an increasingly globalised world. Today, there are pre-trained models (PTMs) specifically designed for automatic machine translation. This paper evaluates the performance of various pre-trained models (PTMs) for many-to-English translation. The models assessed include OPUS-MT \cite{tiedemann-2020-opus-mt}, mBART-50 \cite{liu-2020-mbart}, NLLB-200 \cite{nllb200-2020}, and M2M-100 \cite{fan-2020-m2m100}. Using a dataset curated from the Tatoeba repository, comprising 1,323 parallel sentence pairs across 14 source languages, translation quality is measured using BLEU, SacreBLEU, and METEOR scores. The results indicate that OPUS-MT, a one-to-one PTM, outperforms multilingual models in one-to-English translation tasks, with mBART-50 leading among the multilingual PTMs. Chinese, Italian, Polish, and Turkish translations to English show the best performance on average across the four models used in this study, however, Ukrainian-to-English on the OPUS-MT model achieves the highest SacreBLEU and METEOR scores.
\end{abstract}

@@

This study will lean towards evaluating MT-specific models rather than general LLMs. Through simple experimentations, this paper aims to evaluate the performance of existing and publicly available MT pre-trained models (PTMs) on many-to-English translation across 14 source languages. Although fine-tuning multilingual PTMs has been proven to increase translation performance \cite{cooper-stickland-2021-recipes}, no pre-training or fine-tuning is performed in this study due to simplicity reasons and resource limitations. A dataset is curated from the Tatoeba repository \cite{tatoeba}, containing 1,323 parallel sentence pairs across source and target languages. The models include a one-to-one PTMs: OPUS-MT \cite{tiedemann-2023-democratizing,tiedemann-2020-opus-mt}, and multilingual PTMs: such as mBART-50 \cite{liu-2020-mbart}, NLLB-200 \cite{nllb200-2020}, and M2M-100 \cite{fan-2020-m2m100}. The performance of these models is evaluated using the BLEU \cite{papieni-2002-bleu}, SacreBLEU \cite{post-2018-sacrebleu}, and METEOR \cite{lavie-2007-meteor}.

The result indicates that OPUS-MT outperforms all other multilingual models in one-to-English translation on both SacreBLEU and METEOR scores, highlighting the superiority of specific languages PTMs compared to unfine-tuned multilingual PTMs. Among the multilingual models, mBART-50 shows the best performance, while NLLB-200 performs the worst, by a significant margin. Ukrainian-to-English translation by OPUS-MT receives the highest scores, however, Chinese, Italian, Polish, and Turkish are better on average across the four models.

@@

Currently, pre-trained models (PTMs) have become a cornerstone in machine translation, leveraging large datasets and powerful architectures to achieve state-of-the-art performance \cite{han-2021-ptms}. Attention mechanism has been used for Neural Machine Translation (NMT) \cite{bahdanau-2016-nmt-jointly, luong-2015-effective-attention} even before the introduction of the transformer \cite{vaswani-2017-attention}. The transformer, with its encoder-decoder approach \cite{cho-2014-properties}, remains the foundational architecture for many sequence-to-sequence models in machine translation. Most PTMs support two kinds of translations: one-to-one translation and many-to-many translation (multilingual models).

One-to-one translation refers to a translation approach where a model is specifically trained to translate between one source language and one target language. This setup is characterised by having a dedicated model for each unique language pair. OPUS-MT \cite{tiedemann-2020-opus-mt} is a project dedicated to creating and providing free resources for machine translation, providing over a thousand PTMs of direct one-to-one translations. Additionally, Facebook Fairseq \cite{ott-2019-fairseq} similarly offers one-to-one PTMs, although they are only limited to translations between English, French, German, and Russian.

@@

@@

Table \ref{table:eval_languages} shows the 14 language pairs selected for this study. Languages are chosen based on the resources' availability in Tatoeba and supported languages in most PTMs. Thus, the languages chosen here can be considered moderate to high-resource languages.

@@

OPUS-MT \cite{tiedemann-2023-democratizing,tiedemann-2020-opus-mt} provides over 1,000 pre-trained models for translation between numerous language pairs. The architecture is based on MARIAN-NMT \cite{mariannmt}, based on a standard transformer setup: 6 self-attentive layers in both the encoder and decoder networks, each with 8 attention heads per layer \cite{tiedemann-2020-opus-mt}.

While the framework provides both monolingual and multilingual PTMs, only the OPUS-MT monolingual models are used in this study. This is partly due to the lack of documentation on the multilingual models.

@@

OPUS-MT models are taken from the Helsinki-NLP repository with the exception of 'tiedeman/opus-mt-he-en' for Hebrew-to-English translation. The 'mbart-large-50-many-to-many-mmt', the 'm2m100\_418M', and the 'nllb-200-distilled-600M' models are used for mBART-50 \cite{liu-2020-mbart}, M2M-100 \cite{fan-2020-m2m100}, and NLLB-200 \cite{nllb200-2020}, respectively. All models are taken using the transformers \cite{wolf-etal-2020-transformers} library from Hugging Face \cite{huggingface}. A batch size of 4 is used in all models. Translations are first generated and saved into CSV files, before calculating the metrics score.

For BLEU and METEOR, tokenisation and metrics calculation use the NLTK \cite{bird-2009-natural} package, specifically the \texttt{nltk.word\_tokenize}, \texttt{nltk.translate.meteor\_score}, and \texttt{nltk.translate.bleu\_score.corpus\_bleu}. Additionally, smoothing method 7 is used for the BLEU calculation, which interpolates two other smoothing functions: smoothing method 4, which divides by \( \frac{1}{\ln(\text{len}(T))} \), where \( T \) is the length of the translation; and smoothing method 5, which averages the \( n-1 \), \( n \), and \( n+1 \) gram matched counts \cite{nltk-bleu-docs}. The \texttt{sacrebleu} \cite{post-2018-sacrebleu} package from pip is used to calculate SacreBLEU \cite{post-2018-sacrebleu}.

>>>>

\title{Evaluation of Machine Translation Pre-Trained Models for Many-to-English Translation}
\author{Stefan Liemawan Adji}
\date{\today}

@@

\begin{abstract}
    Machine translation (MT) has become an essential tool for overcoming language barriers in an increasingly globalised world. Today, there are pre-trained models specifically designed for automatic machine translation. This paper evaluates the performance of various machine translation pre-trained models (MT-PTMs) for many-to-English translation. The models assessed include OPUS-MT \cite{tiedemann-2020-opus-mt}, mBART-50 \cite{liu-2020-mbart}, NLLB-200 \cite{nllb200-2020}, and M2M-100 \cite{fan-2020-m2m100}. Using a dataset curated from the Tatoeba repository, comprising 1,323 parallel sentence pairs across 14 source languages, translation quality is measured using BLEU, SacreBLEU, and METEOR scores. The results indicate that OPUS-MT, a one-to-one PTM, outperforms multilingual models in one-to-English translation tasks, with mBART-50 leading among the multilingual models. Chinese, Italian, Polish, and Turkish translations to English show the best performance on average across the four models used in this study, however, Ukrainian-to-English on the OPUS-MT model achieves the highest SacreBLEU and METEOR scores.
\end{abstract}

@@

This study will lean towards evaluating MT-specific models rather than general LLMs. Through simple experimentations, this paper aims to evaluate the performance of existing and publicly available MT pre-trained models (MT-PTMs) on many-to-English translation across 14 source languages. Although fine-tuning multilingual MT-PTMs has been proven to increase translation performance \cite{cooper-stickland-2021-recipes}, no pre-training or fine-tuning is performed in this study due to simplicity reasons and resource limitations. A dataset is curated from the Tatoeba repository \cite{tatoeba}, containing 1,323 parallel sentence pairs across source and target languages. The models include a one-to-one PTMs: OPUS-MT \cite{tiedemann-2023-democratizing,tiedemann-2020-opus-mt}, and multilingual MT-PTMs: such as mBART-50 \cite{liu-2020-mbart}, NLLB-200 \cite{nllb200-2020}, and M2M-100 \cite{fan-2020-m2m100}. The performance of these models is evaluated using the BLEU \cite{papieni-2002-bleu}, SacreBLEU \cite{post-2018-sacrebleu}, and METEOR \cite{lavie-2007-meteor}.

The result indicates that OPUS-MT outperforms all other multilingual models in one-to-English translation on both SacreBLEU and METEOR scores, highlighting the superiority of specific languages MT-PTMs compared to unfine-tuned multilingual MT-PTMs. Among the multilingual models, mBART-50 shows the best performance, while NLLB-200 performs the worst, by a significant margin. Ukrainian-to-English translation by OPUS-MT receives the highest scores, however, Chinese, Italian, Polish, and Turkish are better on average across the four models.

@@

Attention mechanism has been used for Neural Machine Translation (NMT) \cite{bahdanau-2016-nmt-jointly, luong-2015-effective-attention} even before the introduction of the transformer \cite{vaswani-2017-attention}. The Transformer, with its encoder-decoder architecture \cite{vaswani-2017-attention}, has since become the foundational framework for many sequence-to-sequence models in machine translation \cite{cho-2014-properties}. By combining this architecture with training on large datasets, models were able to capture and retain vast amounts of linguistic information, leading to the development of pre-trained models.

Currently, pre-trained models have become a cornerstone in machine translation, achieving state-of-the-art performances \cite{han-2021-ptms}. While general Large Language Models (LLMs) such as BERT \cite{devlin-2019-bert}, GPT \cite{openai-2024-gpt4}, or LLAMA \cite{touvron-2023-llama2} can be used for machine translation tasks, in this study the focus will be towards MT-specific PTMs (MT-PTMs) which are specifically designed, built, and trained for machine translation tasks. Most MT-PTMs typically support two kinds of translations: one-to-one translation and many-to-many translation (multilingual models).

One-to-one translation refers to a translation approach where a model is specifically trained to translate between one source language and one target language. This setup is characterised by having a dedicated model for each unique language pair. OPUS-MT \cite{tiedemann-2020-opus-mt} is a project dedicated to creating and providing free resources for machine translation, providing over a thousand MT-PTMs of direct one-to-one translations. Additionally, Facebook Fairseq \cite{ott-2019-fairseq} similarly offers one-to-one MT-PTMs, although they are only limited to translations between English, French, German, and Russian.

@@

With a lot of

@@

Table \ref{table:eval_languages} shows the 14 language pairs selected for this study. Languages are chosen based on the resources' availability in Tatoeba and supported languages in most MT-PTMs. Thus, the languages chosen here can be considered moderate to high-resource languages.

@@

OPUS-MT \cite{tiedemann-2023-democratizing,tiedemann-2020-opus-mt} provides over 1,000 pre-trained models for translation between numerous language pairs. The architecture is based on MARIAN-NMT \cite{mariannmt}, based on a standard transformer setup: 6 self-attentive layers in both the encoder and decoder networks, each with 8 attention heads per layer \cite{tiedemann-2020-opus-mt}. The models were trained for up to 72 hours on 1 or 4 GPUs, noting that not all of them fully converged within this time frame.

While the framework provides both monolingual and multilingual models, only the OPUS-MT monolingual models are used in this study. This is partly due to the lack of documentation on the multilingual models.

@@

OPUS-MT models are taken from the Helsinki-NLP repository through the Hugging Face website \cite{huggingface,zhang-2023-fine-tuning} with the exception of Hebrew-to-English models taken directly from 'tiedeman/opus-mt-he-en'. The 'mbart-large-50-many-to-many-mmt', the 'm2m100\_418M', and the 'nllb-200-distilled-600M' model variants are used for mBART-50 \cite{liu-2020-mbart}, M2M-100 \cite{fan-2020-m2m100}, and NLLB-200 \cite{nllb200-2020}, respectively. Similarly, all multilingual models are downloaded using the transformers \cite{wolf-2020-transformers} library from Hugging Face \cite{huggingface}. A batch size of 4 is used in all models. Translations from each source language to english are first generated through inference and saved into CSV files, before calculating the metrics score.

For BLEU and METEOR, tokenisation and metrics calculation use the NLTK \cite{bird-2009-natural} package, specifically:
\begin{itemize}
    \item \texttt{nltk.word\_tokenize} \\
    \item \texttt{nltk.translate.meteor\_score} \\
    \item \texttt{nltk.translate.bleu\_score.corpus\_bleu}
\end{itemize}
Additionally, the smoothing method 7 is used for the BLEU calculation, which interpolates two other smoothing functions: smoothing method 4, which divides by \( \frac{1}{\ln(\text{len}(T))} \), where \( T \) is the length of the translation; and smoothing method 5, which averages the \( n-1 \), \( n \), and \( n+1 \) gram matched counts \cite{nltk-bleu-docs}. The \texttt{sacrebleu} \cite{post-2018-sacrebleu} package from pip is used to calculate SacreBLEU \cite{post-2018-sacrebleu}.
